---
title: "BST 270 Final Project"
author: "Nick Birk"
date: "1/17/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The aim of the following project is to recreate one table and one figure from the fivethirtyeight article:

["Inside The Political Donation History Of Wealthy Sports Owners"](https://fivethirtyeight.com/features/inside-the-political-donation-history-of-wealthy-sports-owners/)


# Preparation

To proceed with the analysis, we will first need to load any necessary packages and import the dataset from online. If the dataset is available locally, this option will instead be used to import the dataset.

```{r load_package_and_data, message = FALSE}
# Install packages if they are missing, otherwise load
if (!require('tidyverse')) install.packages('tidyverse')
if (!require('ggplot2')) install.packages('ggplot2')
if (!require('RCurl')) install.packages('RCurl')
if (!require('knitr')) install.packages('knitr')

# Import dataset locally, if available
if (file.exists("../data/sports-political-donations.csv")){
    sports <- read.csv("../data/sports-political-donations.csv")
} else {
    # Otherwise, import dataset from raw github file online
   sports_url <- getURL("https://raw.githubusercontent.com/fivethirtyeight/data/master/sports-political-donations/sports-political-donations.csv")
   sports <- read.csv(text = sports_url)
}


```


# Part 1: Figure 1

To begin, we attempt to recreate the first figure that appears in the article, titled "Team owners give largely to the GOP".

```{r first_inspect}
kable(head(sports))
```

From an inspection of the data, one can see that some processing is required to obtain the values needed to reproduce the figure. In particular, we will first need to process the `donate_amount` colunn from a string (including symbols such as $ and ,) to a numeric value.

```{r clean_amount}
# Process the donation amount by removing $ and , symbols
sports$donate_amount <- str_replace(sports$Amount, "\\$", "")
sports$donate_amount <- str_replace(sports$donate_amount, "\\,", "")
# Second line needed for amounts which originally included multiple "," symbols
sports$donate_amount <- str_replace(sports$donate_amount, "\\,", "")
# Convert from character to numeric
sports$donate_amount <- as.numeric(sports$donate_amount)
```

Now that the donation amount has been made numeric (as the column `donate_amount`), we can proceed. To reproduce the figure, we will want to create a summary within each election year and party. We will use the `group_by` function to get the total donations for each party-election year pair. We will then display the table to ensure we have the correct categories.

```{r summarize_fig1}
# Need to summarize by party, year to get figure values
sports_summary <- sports %>% group_by(Election.Year, Party) %>%
  summarize(total_donated = sum(donate_amount))

# We can see from inspecting this summary table that there are some extra categories
kable(sports_summary)
```

We can see from the above that there are some extra political parties that do not appear in the figure itself. Further, it may be helpful to take an additional step to obtain the proportions of donations for the given year, rather than the raw amount. We make these final processing steps below and display the resulting table:

```{r summarize_fig1_more}
# Create an even smaller summary dataset that includes the proportion
# Calculated as amount / total
sports_annual_totals <- sports_summary %>% group_by(Election.Year) %>%
  summarize(annual_total = sum(total_donated))
# Join these values as new column and then create proportion column
sports_summary <- inner_join(sports_summary, sports_annual_totals)
sports_summary <- sports_summary %>% mutate(donate_prop = total_donated/annual_total) %>% filter(Party %in% c("Republican", "Democrat"))

kable(sports_summary)
```



# Part 2: Table 2

We now attempt to reproduce the second table that appears in the article, titled "MLB owners have donated the most"


